### 2014-11-08 language.tex
More updates.

Linear Meld (LM) is a forward-chaining linear logic programming
language~\cite{cruz-iclp14} with roots on Datalog~\cite{Ramakrishnan93asurvey}.
Programs consist of a set of \emph{rules} and a \emph{database of facts}.
Rules such as \texttt{a(X), b(Y) -o c(X, Y)} can be read as follows: if fact \texttt{a(X)}
and fact \texttt{b(Y)} exist in the database then \texttt{c(X, Y)} is added to
the database. The expression \texttt{a(X), b(Y)} is called the \emph{body} of
the rule and \texttt{c(X, Y)} is the \emph{head} of the rule.
A fact is a predicate, e.g., \texttt{a}, \texttt{b} or \texttt{c}, and its
associated tuple of values, e.g., the values of \texttt{X} and \texttt{Y}, representing the
arguments. Since LM uses linear logic as its foundation, we distinguish between
\emph{linear} and \emph{persistent facts}. Linear facts are deleted during the
process of deriving rule, while persistent facts are not.
Program execution starts by adding the \emph{axioms} (the initial facts) of the program to the database.
Next, the rules are recursively applied and the database is updated by adding
new facts or deleting facts used during rule derivation.
When no more rules are applicable, the program terminates.

\newcommand{\scare}[1]{``#1''}

LM has been designed for writing programs that operate on graphs. LM partitions
the database by using the first argument of each fact. The first argument is
typed as a \emph{node} and represents a node in the graph. For instance, the
fact \texttt{f(@1, 2)} is stored in node \texttt{@1}, while fact \texttt{p(@2)}
is stored in node \texttt{@2}. LM restricts
the body of every rule to only refer to the same node so that nodes can derive
rules independently. Although the body is restricted, the head of the
rule may refer to any node as long as the node is refered somewhere in the
body. This allows \scare{communication} between nodes during rule derivation, since a
node may \scare{send} a fact to another node.
Rule restrictions in turn make LM implicitly parallel because nodes are able to
compute independently. This makes LM non-deterministic since nodes can be picked
to run in any order, affecting which rules are applied and which facts are
deleted or derived.

In order to fully understand how the language works, we present a very simple
algorithm: the single source shortest path program~(SSSP). Later in the paper, we
add coordination facts to improve the execution of the program.

The SSSP program in Fig.~\ref{code:shortest_path_program} starts (lines 1-3)
with the declaration of the predicates. Predicates specify the kinds of facts
used in the program. The first predicate, \texttt{edge}, is a persistent
predicate that describes the relationship between the nodes of the graph,
where the third argument represents the weight of the edge (the
\texttt{route} modifier informs the compiler that the \texttt{edge} predicate
determines the structure of the graph).
The predicates \texttt{shortest} and \texttt{relax} are
specified as linear facts and thus are deleted when deriving new facts.  The
algorithm computes the shortest distance from node
\texttt{@1} to all other nodes in the graph. Every node has a
\texttt{shortest} fact that is improved with new \texttt{relax} facts.  Lines
5-9 declare the axioms of the program: \texttt{edge} facts describe the
graph; \texttt{shortest(A, +00, [])} is the initial shortest distance
(infinity) for all nodes; and \texttt{relax(@1, 0, [@1])} starts the
algorithm with the initial distance from \texttt{@1} to \texttt{@1}.

@@

The first rule of the program (lines 11-14) reads as following: if the current
\texttt{shortest} path \texttt{P1} with distance \texttt{D1} is larger
than a new path \texttt{relax} with distance \texttt{D2}, then replace the
current shortest path with \texttt{D2}, delete the new \texttt{relax} path and
propagate new paths to the neighbors (lines 14-15) using a \emph{comprehension}.
The comprehension iterates over the edges of node \texttt{A} and derives a new
\texttt{relax} fact for each node \texttt{B} with the distance \texttt{D2 + W},
where \texttt{W} is the weight of the edge. For
instance, in Fig.~\ref{fig:shortest_path_program}~(a) we apply rule 1 in node
\texttt{@1} where two new \texttt{relax} facts are derived at node \texttt{@2}
and \texttt{@3}. Fig.~\ref{fig:shortest_path_program}~(b) is the result after
applying the same rule but at node \texttt{2}.

@@

There many opportunities for concurrency in the SSSP program. For instance,
after applying rule 1 in Fig.~\ref{fig:shortest_path_program}~(a), it is
possible to either apply rules in either node \texttt{@2} or node
\texttt{@3}. In our implementation, we partition the graph into subgraphs that
are processed by multiple threads of execution. Eventually, it is no longer to
possible to apply rules and the final result present in
Fig.~\ref{fig:shortest_path_program}~(c) is achieved.

%On the other hand, LM has no natural matching of data and computation to workers (processes, threads),
%since nodes are a program abstraction and part of the program's logic.
%We view the set of nodes as a graph data structure where workers will perform work.
%A worker is able to process any node, although a node cannot be computed by more than one worker
%at the same time. This disallows the manipulation of a node by multiple workers.

>>>>

Coordinated Linear Meld (CLM) is a forward-chaining linear logic programming
language based on LM~\cite{cruz-iclp14}.
Programs consist of a set of \emph{rules} and a \emph{database of facts}.
Rules such as \texttt{a(X), b(Y) -o c(X, Y)} can be read as follows: if fact \texttt{a(X)}
and fact \texttt{b(Y)} exist in the database then \texttt{c(X, Y)} is added to
the database. The expression \texttt{a(X), b(Y)} is called the \emph{body} of
the rule and \texttt{c(X, Y)} is the \emph{head} of the rule.
A fact is a predicate, e.g., \texttt{a}, \texttt{b} or \texttt{c}, and its
associated tuple of values, e.g., the concrete values of \texttt{X} and
\texttt{Y}, substitued for the
arguments. Since LM uses linear logic as its foundation, we distinguish between
\emph{linear} and \emph{persistent facts}. Linear facts are deleted during the
process of deriving rule, while persistent facts are not.
Program execution starts by adding the \emph{axioms} (the initial facts) of the program to the database.
Next, the rules are recursively applied and the database is updated by adding
new facts or deleting facts used during rule derivation.
When no more rules are applicable, the program terminates.

CLM has been designed for writing programs that operate on graphs. CLM partitions
the database by using the first argument of each fact. The first argument has
type \emph{node} and represents a node in the graph. To achieve concurrency,
our implementation partitions the database by the first argument, e.g.,
the fact \texttt{f(@1, 2)} is stored in node \texttt{@1}, while
fact \texttt{p(@2)} is stored in node \texttt{@2}. CLM restricts the body of
every rule to facts with the same node so that nodes can derive rules
independently. Although the body is restricted, the head of the rule may refer
to any node as long as that node is refered to somewhere in the body. This allows
\scare{communication} between nodes during rule derivation, since a node may
\scare{send} a fact to another node.  Rule restrictions in turn make CLM
implicitly parallel because nodes are able to compute independently. This makes
CLM non-deterministic since nodes can be picked to run in any order, affecting
which rules are applied and which facts are deleted or derived.

To make these ideas concrete, we present a simple
program: the single source shortest path program~(SSSP). Later in the paper, we
add coordination facts to improve the execution of the program.

The SSSP program in Fig.~\ref{code:shortest_path_program} starts (lines 1-3)
with the declaration of the predicates. Predicates specify the facts
used in the program. The first predicate, \texttt{edge}, is a persistent
predicate that describes the relationship between the nodes of the graph,
where the third argument represents the weight of the edge (the
\texttt{route} modifier informs the compiler that the \texttt{edge} predicate
determines the structure of the graph).
The predicates \texttt{shortest} and \texttt{relax} are
specified as linear facts and thus are deleted when deriving new facts.  The
algorithm computes the shortest distance from node
\texttt{@1} to all other nodes in the graph. Every node has a
\texttt{shortest} fact that is improved with new \texttt{relax} facts.  Lines
5-9 declare the axioms of the program: \texttt{edge} facts describe the
graph; \texttt{shortest(A, +00, [])} is the initial shortest distance
(infinity) for all nodes; and \texttt{relax(@1, 0, [@1])} starts the
algorithm by setting the initial distance from \texttt{@1} to \texttt{@1} to be
0.

@@

The first rule of the program (lines 11-14) reads as following: if the current
\texttt{shortest} path \texttt{P1} with distance \texttt{D1} is larger
than a new path \texttt{relax} with distance \texttt{D2}, then replace the
current shortest path with \texttt{D2}, delete the new \texttt{relax} path and
propagate new paths to the neighbors (lines 14-15) using a \emph{comprehension}.
The comprehension iterates over the edges of node \texttt{A} and derives a new
\texttt{relax} fact for each node \texttt{B} with the distance \texttt{D2 + W},
where \texttt{W} is the weight of the edge. For
example, in Fig.~\ref{fig:shortest_path_program}~(a) we apply rule 1 in node
\texttt{@1} where two new \texttt{relax} facts are derived at node \texttt{@2}
and \texttt{@3}. Fig.~\ref{fig:shortest_path_program}~(b) is the result after
applying the same rule but at node \texttt{2}.

@@

There are many opportunities for concurrency in the SSSP program. For instance,
after applying rule 1 in Fig.~\ref{fig:shortest_path_program}~(a), it is
possible to either apply rules in either node \texttt{@2} or node
\texttt{@3}. This decision depends largely on implementation factors such as node
partitioning and number of threads in the system.
Still, it is easy to prove that no matter the scheduling used,
the final result present in Fig.~\ref{fig:shortest_path_program}~(c) is achieved.

%On the other hand, CLM has no natural matching of data and computation to workers (processes, threads),
%since nodes are a program abstraction and part of the program's logic.
%We view the set of nodes as a graph data structure where workers will perform work.
%A worker is able to process any node, although a node cannot be computed by more than one worker
%at the same time. This disallows the manipulation of a node by multiple workers.

### 2014-11-08 minimax.tex
More updates.

The algorithm proceeds by building a game tree, where each tree node represents a
game state and the children represent the possible game moves that can
be made by either player 1 or player 2. The
tree levels are built in a way that each player plays in turns and the
first level represents the moves made by player 1.  An
evaluation function \texttt{f(State)} is used to compute the score of
the board for each leaf of the tree. A node is a leaf when the game state can no longer be
expanded. Finally, the
algorithm recursively minimizes or maximizes the scores of each node.
To select the best move for player 1, the
algorithm picks the move maximized at the root node.

@@

The LM code in Fig.~\ref{minimax:check-end} implements
the initial expansion of the tree. The first three rules (lines 1-10) deal
with the cases where no children nodes are created and the last three rules
(12-29) deal with cases where new nodes may be created.
The rule in lines 12-25 generate new nodes using the
\texttt{exists} language construct, which creates a new node address
\texttt{B}, where a new \texttt{play} fact is created with the new
board. A \texttt{parent(B,A)} fact is derived to allow node \texttt{B} to
communicate with its parent.

As noted before in Section~\ref{sec:fifo}, we use a FIFO ordering for nodes,
meaning that when children are expanded, we first expand the children and
then the children of the children. We note
however, that this is not optimal for this program since the complete tree needs
to be expanded before computing the scores at the leaves. The memory
complexity of the program is $\mathcal{O}(n)$, where $n$ is the number of nodes
in the tree.

@@

With coordination, we set the priority nodes to be the same as the
depth so that the tree is expanded in a depth-first fashion (lines 15 and 22).
The memory complexity of the program then becomes $\mathcal{O}(d t)$, where $d$
is the depth of the tree and $t$ is the number of threads. In
Fig.~\ref{fig:minimax}, we have two threads (squares) and the first one has
expanded the tree on the left side. When a second thread is idle, it steals the
first unexpanded node and expands the tree from that position. Since threads
prioritize deeper nodes, the scores of the first leaves are immediatelly
computed and then sent to the parent node. At this point, the leaves are deleted
and reused for other nodes in the tree, resulting in minimal memory usage.

\begin{figure}[h!]
   \begin{center}
      \includegraphics[width=4.5cm]{figures/minimax_tree}
   \end{center}
   \caption{Expanding the MiniMax tree using coordination. By prioritizing
      deeper nodes, threads are forced to expand the tree using a depth-first
      approach, which is superior since there is no need to expand the whole
      tree before computing the node scores.}
   \label{fig:minimax}
\end{figure}

@@

In Fig.~\ref{results:memory_minmax} we compare the memory usage and scalability
of the coordinated MiniMax against the regular MiniMax. The coordinated version
sees a huge reduction of memory usage, needing at most 8MB for 16 threads, while
the regular version needs around 14GB. Note that as the number of threads goes
up, memory usage also goes up. This is an artifact of our parallel memory
allocator that allocates large chunks of memory beforehand.
In terms of run time, our experimental results show a 14-fold speedup for the
coordinated version against a 12-fold for the regular version when using 16
threads. When comparing the two versions directly, there is a 20\% run time
reduction when using 16 threads in the coordinated version.

>>>>

The algorithm proceeds by building a game tree, where each tree node represents a
game state and the children represent the possible game moves that can
be made by either player 1 or player 2.
An evaluation function \texttt{f(State)} is used to compute the score of
the board for each leaf of the tree. A node is a leaf when the game state can no longer be
expanded. Finally, the
algorithm recursively minimizes or maximizes the scores of each node.
To select the best move for player 1, the
algorithm picks the move maximized at the root node.

@@

The code in Fig.~\ref{minimax:check-end} shows part of the program.
The first three rules (lines 1-10) deal
with the case where no children nodes are created and the last three rules
(12-29) deal with the cases for creating new nodes.
The rule in lines 12-25 generates new nodes using the
\texttt{exists} language construct, which creates a new node address
\texttt{B}, where a new \texttt{play} fact is created with the a
board. A \texttt{parent(B,A)} fact is derived to allow node \texttt{B} to
communicate with its parent.

As noted before in Section~\ref{sec:fifo}, the runtime schedules nodes by XXX,
in a FIFO ordering for nodes,
meaning that when children are expanded, we first expand the children and
then the children of the children. In this case, the defined ordering
is not optimal for since the complete tree needs
to be expanded before computing the scores at the leaves. The memory
complexity of the program is $\mathcal{O}(n)$, where $n$ is the number of nodes
in the tree.

@@

With coordination, we set the priority nodes to be the same as the
depth so that the tree is expanded in a depth-first fashion (lines 15 and 22).
The memory complexity of the program then becomes $\mathcal{O}(d t)$, where $d$
is the depth of the tree and $t$ is the number of threads. As an example,
consider a system with 2 threads, $T_1$ and $T_2$, where $T_1$ first expands the root
node and then the first child. Since $T_2$ is idle, it steals half of the root's
children nodes and starts expanding one of the nodes in a depth-first fashion.
Since threads prioritize deeper nodes, the scores of the first leaves are immediatelly
computed and then sent to the parent node. At this point, the leaves are deleted
and reused for other nodes in the tree, resulting in minimal memory usage.

\iffalse
\begin{figure}[h!]
   \begin{center}
      \includegraphics[width=4.5cm]{figures/minimax_tree}
   \end{center}
   \caption{Expanding the MiniMax tree using coordination. By prioritizing
      deeper nodes, threads are forced to expand the tree using a depth-first
      approach, which is superior since there is no need to expand the whole
      tree before computing the node scores.}
   \label{fig:minimax}
\end{figure}
\fi

@@

In Fig.~\ref{results:memory_minmax} we compare the memory usage and scalability
of the coordinated MiniMax against the regular MiniMax. The coordinated version
sees a huge reduction of memory usage, needing at most 8MB for 16 threads, while
the regular version needs almost 14GB. Note that as the number of threads goes
up, memory usage also goes up. This is an artifact of our parallel memory
allocator that allocates large chunks of memory beforehand.
In terms of scalability, our experimental results show a 14-fold speedup for the
coordinated version against a 12-fold for the regular version when using 16
threads. When comparing the two versions directly, there is a 20\% run time
reduction when using 16 threads in the coordinated version.

### 2014-11-08 related.tex
More updates.

There has been an increasing interest in declarative and data-centric
languages. MapReduce~\cite{Dean:2008:MSD:1327452.1327492}, for instance, is a
popular data-centric programming model that is optimized for large clusters. The
scheduling and data sharing model is very simple: in the \emph{map phase}, data
is transformed at each node and the result reduced to a final result in the
\emph{reduce phase}. In order to facilitate the writing of programs over large
datasets, SQL-like languages such as
PigLatin~\cite{Olston:2008:PLN:1376616.1376726} have been developed.
PigLatin builds on top of MapReduce and allows the programmer to write complex
data-flow graphs, increasing the abstraction of the MapReduce paradigm.
An alternative to PigLatin/MapReduce is
Dryad~\cite{Isard:2007:DDD:1272996.1273005}
that allows programmers to design arbitrary computation patterns using the DAG
abstraction. It combines computational vertices with communication channels
(edges). Dryad programs are automatically scheduled to run on multiple
computers/cores and data is partitioned automatically during runtime.

@@

Linda and Delirium are limited in the sense that the programmer can only
coordinate the scheduling of processing units, while the placement of data is
left to the implementation. LM differs from those languages since it
is fully data-driven and coordination acts on data instead of processing units.
The abstraction is then raised by considering data and algorithmic aspects of
the program instead of focusing how processing units are used.
Furthermore, the LM language is both a coordination language and a computation
language and there is no distinction between the two components.

Forward-chaining logic programming was made famous by the Datalog
language~\cite{Ullman:1990:PDK:533142}.  Traditionally used in deductive
databases, it is now being increasingly used in different fields such as
distributed networking~\cite{Loo-condie-garofalakis-p2}, sensor
nets~\cite{Chu:2007:DID:1322263.1322281} and cloud computing~\cite{alvaro:boom}.
LM itself is inspired in Meld~\cite{ashley-rollman-iclp09}, a Datalog-like language for
programming distributed emsembles of modular robots. Meld introduced the idea of
sensing and action facts in order to sense and act on the outside world,
respectively. Coordination facts are inspired in the same principles but are
used for coordinating data and computation.

>>>>

There has been an increasing interest in declarative 
languages. MapReduce~\cite{Dean:2008:MSD:1327452.1327492}, for instance, is a
popular data-centric programming model that is optimized for large clusters.
Intrinsic to its popularity is the simplicity of its scheduling and data sharing model.
In order to facilitate the writing of programs over large
datasets, SQL-like languages such as
PigLatin~\cite{Olston:2008:PLN:1376616.1376726} have been developed.
PigLatin builds on top of MapReduce and allows the programmer to write complex
data-flow graphs, raising the abstraction and easy of programmability of
MapReduce programs.
An alternative to PigLatin/MapReduce is
Dryad~\cite{Isard:2007:DDD:1272996.1273005}
that allows programmers to design arbitrary computation patterns using the DAG
abstraction. It combines computational vertices with communication channels
(edges). Dryad programs are automatically scheduled to run on multiple
computers/cores and data is partitioned automatically during runtime.

@@

Linda and Delirium are limited in the sense that the programmer can only
coordinate the scheduling of processing units, while the placement of data is
left to the implementation. But even more importantly, CLM differs from those
languages because coordination acts on data instead of processing units.
The abstraction is then raised by considering data and algorithmic aspects of
the program instead of focusing how processing units are used.
Furthermore, the LM language is both a coordination language and a computation
language and there is no distinction between the two components.

Forward-chaining logic programming came into its own with the Datalog
language~\cite{Ullman:1990:PDK:533142}.  Traditionally used in deductive
databases, it is now being increasingly used in different fields such as
distributed networking~\cite{Loo-condie-garofalakis-p2}, sensor
nets~\cite{Chu:2007:DID:1322263.1322281} and cloud computing~\cite{alvaro:boom}.
The LM language was inspired by Meld~\cite{ashley-rollman-iclp09}, a Datalog-like language for
programming distributed emsembles of modular robots. Meld introduced the idea of
sensing and action facts in order to sense and act on the outside world,
respectively. LCM uses sensing facts to sense the underlying system that the program is running on and action facs
act on that system.

### 2014-11-08 sssp.tex
More updates.

The Single Source Shortest Path (SSSP) program was already presented in
Section~\ref{sec:language}. We are going to update the code in
Fig.~\ref{code:shortest_path_program} so that nodes with the shortest distances
are selected to run before others nodes. The coordinated version of the same
algorithm is presented in Fig.~\ref{code:shortest_path_program_coord} and it
uses one coordination fact, namely \texttt{set-priority} (line 14). Note that we
also use a global program directive to order priorities in ascending order (line
5).

When using a single thread, the algorithm behaves like
Dijkstra's shortest path algorithm~\cite{Dijkstra}. However, when using multiple
threads, each thread will pick the smallest distance from their subset of nodes.
While this is not optimal since threads do not share a global view of priorities,
many rule derivations are going to be avoided.

@@

Figure~\ref{results:sssp_uspowergrid} shows experimental results for a variant of
the SSSP program that computes the SSSP for many of the nodes of the graph.
There are some situations where unecessary facts are propagated
because although the shortest distance is selected, sub-optimal distances may be
propagated because many SSSP distances are computed at the same time.
However, we see a reduction of over 50\% in the number of
derived facts when using the coordinated version \textbf{C} over the regular
version \textbf{R}. The results also show that the cordinated
version using 16 threads is 1.3 times faster than the regular version using 16
threads. In turn, this makes the coordinated version 20 times faster
than the regular version using 1 thread. Note that the coordinated program is as
scalable as the regular program.

\begin{figure}[h!]
   \begin{center}
      \subfloat[]{ \begin{tabular}[b]{ | c | c | c | }
         \hline                       
         \textbf{\# T} & \textbf{R} & \textbf{C} \\ \hline \hline
         1 & 333K & 206K \\
         2 & 300K & 210K \\
         4 & 316K & 208K \\
         8 & 328K & 211K \\
         16 & 343K & 212K \\
         \hline  
         \end{tabular}
         \normalsize
      }
      \subfloat[]{\includegraphics[width=4.5cm]{results/shortest-uspowergrid.png}}
   \end{center}
   \caption{Experimental results for the SSSP program using US powergrid network.
   On the left we show the number of facts derived for the regular \textbf{R}
   and coordinated version \textbf{C} using a variable number of threads
   \textbf{\# T}. On the right, we have the scalability of the regular and
   coordinated version. The speedup values are computed using the execution
   time for 1 thread.}
   \label{results:sssp_uspowergrid}
\end{figure}

>>>>

In Section~\ref{sec:language}, we said that the SSSP program has an optimal
scheduling if the nodes with the shortest distances are selected to run before
other nodes.  The coordinated version of the SSSP
program is presented in Fig.~\ref{code:shortest_path_program_coord} and it
uses one coordination fact, namely \texttt{set-priority} (line 14). Note that we
also use a global program directive to order priorities in ascending order (line
5).

When using a single thread, the algorithm behaves like
Dijkstra's shortest path algorithm~\cite{Dijkstra}. However, when using multiple
threads, each thread will pick the smallest distance from their subset of nodes.
While this does not yield the optimal program with relation to 1 thread, it
allows for parallel execution and locally avoid unnecessary work. XXX

@@

Figure~\ref{results:sssp_uspowergrid} shows experimental results a
SSSP program that computes the SSSP for 20\% of the nodes of the graph.
There are some situations where unecessary facts are propagated
because although the shortest distance is selected, sub-optimal distances may be
propagated because many SSSP distances are computed at the same time.
However, we see a reduction of over 50\% in the number of
derived facts when using the coordinated version \textbf{C} over the regular
version \textbf{R}. The results also show that the cordinated
version using 16 threads is 1.3 times faster than the regular version using 16
threads. In turn, this makes the coordinated version 20 times faster
than the regular version using 1 thread. Note that the coordinated program is as
scalable as the regular program.

\begin{figure}[h!]
   \begin{center}
      \subfloat[]{ \begin{tabular}[b]{ | c | c | c | }
         \hline                       
         \textbf{\# T} & \textbf{R} & \textbf{C} \\ \hline \hline
         1 & 333K & 206K \\
         2 & 300K & 210K \\
         4 & 316K & 208K \\
         8 & 328K & 211K \\
         16 & 343K & 212K \\
         \hline  
         \end{tabular}
         \normalsize
      }
      \subfloat[]{\includegraphics[width=5.5cm]{results/shortest-uspowergrid.png}}
   \end{center}
   \caption{Experimental results for the SSSP program using US powergrid network.
   On the left we show the number of facts derived for the regular \textbf{R}
   and coordinated version \textbf{C} using a variable number of threads
   \textbf{\# T}. On the right, we have the scalability of the regular and
   coordinated version. The speedup values are computed using the execution
   time for 1 thread.}
   \label{results:sssp_uspowergrid}
\end{figure}
